---
title: Enhance ETL pipeline monitoring with text plots
author: Rahul
date: '2021-07-19'
slug: text-plots
categories:
  - R
tags:
  - visualization
  - docker
output: 
  blogdown::html_page:
    toc: true
---

```{r include=FALSE}
library(dplyr)
library(txtplot)
```

I often use data ETL pipeline scripts cron-jobed on text-only interfaces (be it Jenkins, or cron-jobed in shell). While I print  descriptive stats to keep tabs on ETL runs, I've found [`{txtplot}`](https://cran.r-project.org/web/packages/txtplot/index.html) adds a higher level of fidelity in my logfiles. [Bjoern Bornkamp](https://www.rdocumentation.org/collaborators/name/Bjoern%20Bornkamp)'s package offers a simple and effective way to enhance my logfiles. 

Will I use them to visualize my data? No. Are they useful tokeep an eye on your pipelines and quickly diagnose issues? Yes, I've been able to diagnose more than a few data quality spills very quickly because I had these rudimentary plots in my logfiles.

Some examples...

### What's the distribution of a variable?

```{r}
dat <- dplyr::starwars %>% 
  mutate(across(where(is.character), forcats::fct_infreq))

dat %>%
  tidyr::drop_na(height) %>% 
  pull(height) %>% 
  txtdensity(., xlab = "starwars: height distribution", width = 70, height = 12, pch = "o")
```

### Counts of a factor?

```{r}
txtbarchart(dat$sex, width = 70, height = 12, pch = "x")
```

### Boxplots

```{r}
txtboxplot(dat$height, dat$mass, width = 70, height = 12)
```

### Time series plots

```{r}
txtplot(y = LakeHuron, x = 1875:1972, width = 70, height = 12, xlab = "year", ylab = "level")
```

ACF plots, though I've not used these in production yet.

```{r}
txtacf(sunspot.year, width = 70, height = 12)
```

