---
title: Finite Mixture Modeling using Flexmix
author: Rahul Sangole
date: '2017-02-01'
slug: finite-mixture-modeling-using-flexmix
tags:
  - R
  - flexmix
  - mixture modeling
output:
  blogdown::html_page:
    toc: true
---
```{r setup, include=FALSE}
knitr::opts_chunk$set(cache=TRUE)
```

This page replicates the codes written by Grun & Leish (2007) in ['FlexMix: An R package for finite mixture modelling', University of Wollongong, Australia.](http://ro.uow.edu.au/cgi/viewcontent.cgi?article=3410&context=commpapers) My intent here was to learn the flexmix package by replicating the results by the authors.

# Model Based Clustering

The model based clustering on the whiskey dataset. The whiskey dataset is from the Simmons Study of Media and Markets (Fall 1997), and contains the incidence matrix for scotch brands in households who reported consuming scotch for period of 1 year. The dataset is taken from [Edwards and Allenby (2003).](https://www.researchgate.net/profile/Greg_Allenby/publication/247837327_Multivariate_Analysis_of_Multiple_Response_Data/links/547480910cf2778985abe334.pdf)

Load the necessary packges first: `tidyverse` and `flexmix`.
```{r message=FALSE, warning=FALSE}
library(tidyverse)
library(flexmix)
```

## Quick EDA
Quick look at the data itself. The dataframe consists of 2 elements - frequency (numeric vector), and the incidence matrix. There are total of 484 observations.

```{r}
data("whiskey")
df <- whiskey
set.seed(1802)
str(df)
```

The column names of the `df$Incidence` matrix are the brands of whiskey.
```{r}
colnames(df$Incidence)
```

The incidence matrix shows a relationship between two classes of variables - in this case: freqencies of the brand of whiskey in the past year, and the brand of whiskey itself. Quick look at a portion of the matrix:
```{r}
df$Incidence[sample(x = 1:484,size = 10),sample(1:21,3)]
```

The popularity of the whiskeys can be seen here. Chivas Regal seems to be a favourite, which puts my personal preference in line with a larger population :)

```{r, message=FALSE, warning=FALSE}
c <- colSums(df$Incidence)
d1 <- data.frame(Brand=names(c),counts=c,row.names = NULL)
d1 <- d1 %>% left_join(whiskey_brands) %>% arrange(-counts)
ggplot(d1,aes(reorder(Brand,counts),counts,fill=Type))+geom_bar(stat='identity')+coord_flip()+labs(y='Counts',x='Whiskey Brand')
```

## Model building

The first model in the paper is a stepped Flexmix model, specific for binary variables using the `FLXMcvbinary()` model. Since the objective is to cluster the model based on the Incidence counts and Frequencies, the formula used is `Incidence ~ 1`. The frequencies themselves are input as `weights` in the formula.

```{r}
wh_mix <- stepFlexmix(Incidence ~ 1,
                      weights = ~ Freq, 
                      data = df,
                      model = FLXMCmvbinary(truncated = TRUE),
                      control = list(minprior = 0.005),
                      k=1:7,
                      nrep=5)
summary(wh_mix)
```

A top model can be selecting using BIC or AIC criteria. The BIC criteria selects a model with 5 clusters.
```{r}
plot(BIC(wh_mix),type='b',ylab='BIC')
points(x = which.min(BIC(wh_mix)),min(BIC(wh_mix)),col='red',pch=20)
wh_best <- getModel(wh_mix,'BIC')
print(wh_best)
```

The proportions of the observations in each cluster are shown here:
```{r}
round(prop.table(table(wh_best@cluster)),2)
```

The parameter estimates plotted for model with k=5 is shown below graphically. Component 3 (4% of households) contain the largest number of different brands. Component 1 (25% of households) seen to prefer single malt whiskeys. Component 4 (23% of households) are across the board with Brands, but perhaps show lesser of an interest in single malts, just like Component 5 (29% of the households). 

```{r, fig.height=5, fig.width=8, message=FALSE, warning=FALSE}
# wh_best.prior <- prior(wh_best)
wh_best.param <- parameters(wh_best)
wh_best.param <- data.frame(Brand=stringr::str_replace(rownames(wh_best.param),pattern = 'center.',replacement = ''),
wh_best.param,row.names = NULL) 
wh_best.param <- wh_best.param %>% gather(Components,Value,Comp.1:Comp.5)
wh_best.param <- wh_best.param %>% left_join(y = whiskey_brands,by = 'Brand')
ggplot(wh_best.param,aes(y=Value,x=Brand,fill=Type))+
    geom_bar(stat='identity')+
    coord_flip()+
    facet_grid(.~Components)
```

# Mixtures of Regressions

The next example in the paper is the patent data in Wang et al. (1998). The help file `?patent` notes that the data consists of the number of patents, R&D spending and sales in millions of dollar for 70 pharmaceutical and biomedical companies in 1976, taken from the National Bureau of Economic Research R&D Masterfile. 

## Quick EDA

The dependant variable here is `Patents`. Independant variable is `lgRD` which is the log of R&D spending. The objective in this exercise is to try and find how many may clusters exist within this bi-variate dataset. When I started this exercise, it seemed quite moot to me, since visually, I couldn't really tell any distict clusters. But, the results show otherwise.

```{r}
data("patent")
df_patent <- tbl_df(patent)
df_patent
plot(Patents~lgRD,df_patent)
```

## Model Building

The paper mentions that Wang et al. (1998) chose a finite mixture of three Poisson regression models to represent the data. The `FLXMRglm()` is used for the Poisson model with a concomitant variable modeled using `FLXPmultinom()`.

```{r}
pat_mix <- flexmix(Patents ~ lgRD, k = 3, data = df_patent, model = FLXMRglm(family = "poisson"), concomitant = FLXPmultinom(~RDS))
pat_mix
```

The clusters obtained from the analysis are given by a `cluster()` function.

```{r}
clusters(pat_mix)
```

## Results

The data is replotted but with colors for the clusters and additional splines. As we can see, the model beautifully models three lines through three clusters in the data.

```{r}
Components <- factor(clusters(pat_mix))
xyplot(Patents~lgRD,groups = Components,df_patent,type=c('p','spline'))
```

## Further investigation

The `flexmix` package has a function to plot rootograms of the posterior probabilities of observations. Observations where the a-posteriori probability is large for component #1 and #3 are indicated. As we can see where component #1 has highest probabilities indicated in the 1st bucket, they are lowest in #2 and #3 buckets.

```{r}
plot(pat_mix,mark=1)
plot(pat_mix,mark=3)
```

A summary of the mixture model results show the estimated priors, number of observations within each cluster (size), number of observations with p>10^-4 (post>0), and a ratio of the two. The rations of 0.58, 0.42 and 0.18 indicate big overlaps of the clusters. This can also be observed by the large portion of values in the mid-section of the rootogram above. 

```{r}
summary(pat_mix)
```

Tests of significance of the coefficients are obtained by the `refit()`. In each cluster the intercept and lgRD are both statistically significant at the 0.05 level. The black bars in the plot are 95% CI over the point estimates.

```{r}
rm <- refit(pat_mix)
summary(rm)
plot(rm,bycluster=F)
```

The authors note that "estimates vary between all components even though the co- efficients for lgRD are similar for the first and third component".

# Notes

* For both the models, although I'm using the exact datasets and seed values, I observe different values for proportions in each cluster (for model #1) as well as P values for significance etc (for model #2). This could be due to some changes in the underlying flexmix codes since 2007.

# References

* http://ro.uow.edu.au/cgi/viewcontent.cgi?article=3410&context=commpapers
* http://citeseerx.ist.psu.edu/viewdoc/download?doi=10.1.1.491.3320&rep=rep1&type=pdf
* https://cran.r-project.org/web/packages/flexmix/index.html
* https://cran.r-project.org/web/packages/flexmix/vignettes/mixture-regressions.pdf
* https://cran.r-project.org/web/packages/flexmix/vignettes/flexmix-intro.pdf
* https://cran.r-project.org/web/packages/flexmix/vignettes/bootstrapping.pdf
* [https://www.researchgate.net/profile/Greg_Allenby/.../...e334.pdf](https://www.researchgate.net/profile/Greg_Allenby/publication/247837327_Multivariate_Analysis_of_Multiple_Response_Data/links/547480910cf2778985abe334.pdf)

