---
title: Pur(r)ify Your Carets
author: Rahul Sangole
date: '2017-09-17'
slug: pur-r-ify-your-carets
draft: false
categories:
  - R
tags:
  - R
  - purrr
  - caret
---


```{r, include=FALSE}
library(tidyverse)
library(magrittr)
library(purrr)
library(caret)
library(mlbench)
```
### The motivation

I want to write a quick blogpost on two phenomenal pieces of code written by Kuhn et al, and Wickham et al, namely - [purrr](https://cran.r-project.org/web/packages/purrr/index.html), [caret](https://cran.r-project.org/web/packages/caret/index.html). These play so well with the [tidyverse](https://cran.r-project.org/web/packages/tidyverse/index.html) that they have become an indispensible part of my repertoire. 

tl;dr : 

In any datascience project, I want to investigate the effect of various combinations of:

* Variable transformations
* Variable selection
* Grouped vs ungrouped categorical variables
* Models of different types
* Different hyperparameter tuning methods

For each of the various combinations possible, I want to quantify model performance using common performance metrics like AIC or SBC. Commonly, I'll select the model that has the 'best' possible performance among all such models.

Traditionally, I end up with many R objects: one for each new combination of transformation-model_type-tuning_method. For example, `boostFit`, `xgbFit`, `glmFit`, `elastinetFit` for untransformed variables. If I have any transformations, I might also have `boostFit.xform`, `xgbFit.xform`, `glmFit.xform` etc. Add to that, investigation of grouped vs ungrouped variables... `boostFit.xform.grouped`, `xgbFit.xform.ungrouped` etc. You get the idea. 

The challenge with this approach is that the data and the models remain separated, there's a lot of repeat code for object management, manipulation and plotting, and in order to compare all the models together, we have to somehow stitch the results together. (For the last point, `resamples()` in `caret` works beautifully, but requires the same number of resamples in each model.)

The approach I'm presenting below is a combination of a few approaches I learnt through the [APM](http://appliedpredictivemodeling.com/) book, the `caret` documentation,  grammar and verbage in `tidyverse`, as well as a couple of useful talks in the 2017 R Studio conferenence in Orlando [Notably ones on [purrr](https://www.rstudio.com/resources/videos/happy-r-users-purrr/) and  [list-cols](https://www.rstudio.com/resources/videos/using-list-cols-in-your-dataframe/)]. What you'll also see is that the code is extremely succint, which is simply a joy to write and read.

### The objective

### An example

#### Load the libraries & data

The libraries I'm using here are `tidyr`, `dplyr`, `magrittr`, `purrr`, and `caret`. The dataset is from `mlbench`. 

```{r}
library(tidyr)
library(dplyr)
library(magrittr)
library(purrr)
library(caret)
library(mlbench)
data("BostonHousing")
```

#### Create new variables

For the purposes of this demonstration, I'll simply create two new sets variables using a Box-Cox transformation - `caret`'s `preProcess()` makes this easy - and the squared values of the originals. Save each new variable-set in a new character vector which follows the naming convention `predictors.xxxx`. This makes it super easy to find all such variable sets quickly using `ls(pattern = 'predictors')`.

```{r}
# The originals
response <- 'medv'
predictors.original <- colnames(BostonHousing[,1:13])

# Box-Cox transformation
prepTrain <- preProcess(x = BostonHousing[,predictors.original], method = c('BoxCox'))
boxcoxed <- predict(prepTrain,newdata = BostonHousing[,predictors.original])
colnames(boxcoxed) <- paste0(colnames(boxcoxed),'.boxed')
predictors.boxcoxed <- colnames(boxcoxed)

# Squaring
squared <- (BostonHousing[,c(1:3,5:13)])^2
colnames(squared) <- paste0(colnames(squared),'.sq')
predictors.sq <- colnames(squared)

# All together now...
BostonHousing %<>% 
  cbind(boxcoxed,squared)

# Make sure everything is a numerical (for xgboost to work), and also NOT a tibble (some caret functions have trouble with tibbles)
BostonHousing %<>% 
  map_df(.f = ~as.numeric(.x)) %>% as.data.frame()

glimpse(BostonHousing)
```

Here's our new predictor var-sets:
```{r}
columnNames <- ls(pattern = 'predictors')
print(columnNames)
```

#### Create a starter dataframe


```{r}
no_of_varselect <- length(columnNames)
list(BostonHousing) %>% 
    rep(no_of_varselect) %>% 
    enframe(name = 'id', value = 'data') %>% 
    mutate(columnNames = columnNames) -> starter_df

```

```{r}
rpartModel <- function(X, Y) {
    ctrl <- trainControl(
        ## 5-fold CV
        method = "repeatedcv",
        number = 5
    )
    train(
        x = X,
        y = Y,
        method = 'rpart2',
        trControl = ctrl,
        tuneGrid = data.frame(maxdepth=c(2,3,4,5)),
        preProc = c('center', 'scale')
    )
}
xgbTreeModel <- function(X,Y){
    ctrl <- trainControl(
        ## 5-fold CV
        method = "repeatedcv",
        number = 5
    )
    train(
        x=X,
        y=Y,
        method = 'xgbTree',
        trControl = ctrl,
        tuneGrid = expand.grid(nrounds = c(200), 
                              max_depth = c(2) ,
                              eta = c(0.1),
                              gamma = 1, 
                              colsample_bytree = 1, 
                              min_child_weight = 1, 
                              subsample = 1),
        preProc = c('center', 'scale')
    )
}

model_list <- list(rpartModel=rpartModel,
                   xgbModel=xgbTreeModel) %>%
    enframe(name = 'modelName',value = 'model')

model_list
```


```{r}
df.train <-
    starter_df[rep(1:nrow(starter_df),nrow(model_list)),]

df.train <- df.train %>%
    bind_cols(
        model_list[rep(1:nrow(model_list),nrow(starter_df)),] %>% arrange(modelName)
    ) %>%
    mutate(id=1:nrow(.))

```



```{r}
filterColumns <- function(x,y){
    x[,(colnames(x) %in% eval(parse(text=y)))]
}

df.train %<>%
  mutate(
  train.X = map2(data, columnNames,  ~ filterColumns(.x, .y)),
  train.Y = map(data, ~ .x$medv)
  ) %>%
  dplyr::select(-data) %>%
  mutate(params = map2(train.X, train.Y,  ~ list(X = .x, Y = .y)))
  
df.train %<>% 
    mutate(
        modelFits=invoke_map(model,params),
        RMSE=map_dbl(modelFits,~max(.x$results$RMSE)),
        RMSESD=map_dbl(modelFits,~max(.x$results$RMSESD)),
        Rsq=map_dbl(modelFits,~max(.x$results$Rsquared)),
        bestTune=map(modelFits,~.x$bestTune)
    )

df.train
```